---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```

\  
\  

# Notes REX006
Today we will mainly work with finalizing our script. First just a brief
quiz on R fundamentals (a few questions of this type will be included in the 
written exam of Method1).


\  

## R quiz
For each block of code, select the output you would expect. Use R to 
check your answers.

**Q1**
```{r, eval = FALSE}
5 + 4 - 3 * 3
```
1. 18
2. 0
3. 8
4. error message

**Q2**
```{r, eval = FALSE}
a <- c(1, 2, 3.5)
str(a)
```
1. chr [1:3] 1 2 3.5
2. int [1:3] 1 2 3.5
3. num [1:3] 1 2 3.5
4. error message

**Q3**
```{r, eval = FALSE}
a <- 100
b <- 50
c <- 25
test <- (b > c) & !(a < b)
test
```
1. FALSE
2. TRUE
3. NA
4. error message

**Q4**
```{r, eval = FALSE}
a <- 100
b <- c(25, 50, 100)
test <- b > a
test
```
1. FALSE
2. FALSE FALSE FALSE
3. FALSE FALSE TRUE
4. error message

**Q5**
```{r, eval = FALSE}
a <- c('red', 'blue', 'green')
b <- c(25, 50)
d <- data.frame(a, b)
d$a[1]
```
1. 'red'
2. 'red', 'blue', 'green' 
3. 25, 50
4. error message


**Q6**
```{r, eval = FALSE}
a <- c(3, 7, 4, 5, 6)
a[c(3, 5)]
```
1. 3 5
2. 4 6
3. NA
4. error message

**Q7**
```{r, eval = FALSE}
b <- c(3, 7, 4, 5, 6)
b[-2]
```
1. 3 4 5 6
2. -7
3. 7
4. error message

**Q8**
```{r, eval = FALSE}
a <- c(1, 3, 5)
b <- c(2, 1, 6, 3, 10)
b[a]
```
1. 1 3 5
2. 2 6 10
3. NA NA NA
4. error message


**Q9**
```{r, eval = FALSE}
a <- c(1, 3, 5)
b <- c(7, 3, 4)
c <- c(3, 4, 5)
d <- data.frame(a, b, c)
d[ , 2]
```
1. 3 3 4
2. 7 3 4
3. 3
4. error message


**Q10**
```{r, eval = FALSE}
check_order <- function(x, y, z) {
  strict_order <- 1 * ((x < y) & (y < z))
  strict_order
}
check_order(pi, 4, 2^2)
```
1. 1
2. 0
3. FALSE
4. error message

\  
\  

## Script 1: DAG simulation (to be finalized)
Click
<a href="script1b_6.R" target="_blank">here</a>
for Script1b as we left it after the last seminar (with some changes added later on).  



