---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```

\  
\  

# Notes REX003
Today we'll practice 'subsetting', i.e., extracting elements from data structures.
R will make a lot more sense, once you have understood the basics of
subsetting. Then we will continue working on our simulated experiment.

\  


## Subsetting
Please spend time on the examples below; modify and extend them to get a grip of
the various ways to extract parts of R objects. I've heard of people who gave up 
on R because they couldn't find out how to calculate means for separate groups. 
It's not totally intuitive, but once you know how, it's easy. And more
important, subsetting in R is extremely powerful, allowing you to analyze any 
subsets of your data. 

\  

### Vectors
```{r}
z <- seq(1, 100, 7)
z
```

Use square brackets to subset parts of a vector (elements of vectors are 
indexed 1, 2, ..., n).
```{r}
z[2]  # Second element
z[-2]  # All but the second element
z[c(1,7,9)]  # Some selected elements
z[z > 50]  # All elements greater than 50
z[length(z)]  # Last element
z[z < 1]  # Empty vector returned
```

You have to use square brackets, ordinary brackets are for functions.
```{r}
z(3)
```

Here is an example of the power of R: Subsetting one vector based on 
values in another variable.
```{r}
set.seed(123)
# IQs of a lot of people 
iq <- rnorm(10^5, mean = 100, sd = 15) 
# Their gender
gender <- sample(c('male', 'female', 'other'), length(iq),
                 replace = TRUE, prob = c(.49, .49, .02))
# Gender of the smartest!
gender_smartest <- gender[iq > 145]  
table(gender_smartest)
```

\  

### Data frames
A data frame is a collection of vectors of the same length organized in rows
(observations) and columns (variables). Use the dollar sign to extract a
column (i.e., vector) from a data frame. You may then use square brackets to find 
subsets of data within the vector. Below an example 
using one of R's preloaded data sets (use data() to list all of preloaded data sets). 
```{r}
# Assign R's preloaded data set 'sleep' to data frame d. It contains data 
# on the effect of two sleep drugs on the amount of sleep re control nights
# in two groups of 10 patients each. 
d <- sleep   
str(d)  # Check structure of d

# Subset outcome data: extra sleep on drug
d$extra  
# Subset outcome for group 1
d$extra[d$group == 1]  
# Calculate difference between group medians 
median(d$group == 1) - median(d$group == 2)
# Group membership of those with positive effect
d$group[d$extra > 0]
```

You may also use square brackets to subset a data frame, but then
you need two indices: one for rows, and the other for columns, separated by a 
comma. If you leave one empty, R assumes that you mean all rows or all columns.
```{r}
d[2, 1]  # Value of element in the second row of column 1
# Subset all the data for group 1; the output is a new data frame 
g1 <- d[d$group == 1,]  
# Here's the same without dollar sign (as verified below)
g1b <- d[d[,2] == 1,]
identical(g1, g1b)
```

\  

### Exercises
Create a vector called `x` with 1000 random numbers drawn from N(0,1):
`x <- rnorm(1000)`.

1. Create a new vector y with all elements of x greater than 0, and calculate
the difference between the number of elements in y and the the number you would
expect on average if you repeated the procedure many many times.
2. As 1, but for elements in x between -1 and 1. 
3. As 1, but for elements in x smaller than -2 or greater than 2.

Load R's predefined data set `sleep` and use subsetting to create the
following data structures:

4. A data frame only containing data from the patients with positive outcomes
(positive values on the variable `extra`)
5. A numeric vector with all group values (1 or 2). [Hint: `as.numeric()` 
is useful]
6. A data frame with ID-numbers and outcomes for the five patients with largest
gain in extra sleep. 

\  

## Script 1: Simple DAGs (continued)
Click
<a href="script1_3.R" target="_blank">here</a>
for Script1 as we left it after the last seminar.  

Today we will simulate a scenario with a mediating variable

<br>

**Z is a mediator** 

```{r echo = FALSE}
library(dagitty)
dag3a <- dagitty( "dag {
   X -> Z
   Z -> Y
}")

coordinates(dag3a) <-
list( x = c(X = 1, Y = 2, Z = 1.5),
      y = c(X = 2, Y = 2, Z = 2) )

plot(dag3a)
```

<br><br><br>

**Z is a descendent of a mediator** 

```{r echo = FALSE}
library(dagitty)
dag3b <- dagitty( "dag {
   X -> W
   W -> Y
   W -> Z
}")

coordinates(dag3b) <-
list( x = c(X = 1, Y = 2, W = 1.5, Z = 1.5),
      y = c(X = 2, Y = 2, W = 2, Z = 1) )

plot(dag3b)
```



### Exercises
Next time we will simulate the simple scenario of the "Obesity Paradox", please try it yourself!



\  
\  





