---
title: False
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)

```

\  
\  


# Notes REX004
You have already used several R functions. Today we'll take a closer look at 
how they work. After that we'll continue working on the script: it's finally 
time to do some statistical analyses of the result of our imaginary experiment. 
  
\  

## Functions
Functions are small programs that do things. Most functions require that you
specify what you want to be done by providing arguments, but some functions can
be run without arguments: 
```{r}
date()
```

A function's arguments is listed in its help page. Many functions have 
arguments with default values. For example, the `mean()`
function has as default not to excluded missing data: `na.rm = FALSE`. This
is a bit annoying, because R will give you an NA where you expected a number. But 
it is in fact helpful since it reminds you that your data contain missing values. 
```{r}
my_data <- rnorm(1000)
my_data[my_data > 2 ] <- NA  # Turn values > 2 into missing data (NA)
mean(my_data)  # The mean is NA if NAs are included
mean(my_data, na.rm = TRUE)  # This gives the mean after removing NAs
```

If you don't specify what argument your are providing, R will assume 
that you're entering them in the same order as specified by the function. For
example, our favorite `rnorm()` has this order of arguments: 
`rnorm(n, mean = 0, sd = 1)`. 

So if you call `rnorm(3, 100, 15)`, R will assume that you mean n = 3,
mean = 100 and sd = 15 and give you three random numbers from N(100, 15). 
```{r}
set.seed(1)
rnorm(3, 100, 15)
```

Note that you can input them in another order if you specify each argument: 
```{r}
set.seed(1)
rnorm(sd = 15, n = 3, mean = 100)
```
When writing a script, it's generally a good idea to specify arguments also 
when they are provided in the right order. This reminds you and others reading 
your code what the arguments stand for.
  
\  


### Writing your own functions
You can easily create your own functions. Here is an example of a function that
takes a numeric vector as input and returns a z-transformed version of the data. 
```{r}
z_transform <- function(x) {
  zvalue <- (x - mean(x)) / sd(x)
  zvalue  # This returns the output 
}
```
When you run this code, it is stored as an object in R's memory (you should see it 
in R-Studio's Global Environment panel) and will be available until you
close R or clear its workspace with `rm(list = ls())` .
```{r}
my_data <- c(1, 4, 9, 9, 14, 17)
my_z <- z_transform(my_data) 
my_z 
```

Note that the objects created inside a function (e.g., `zvalue` in 
`z_transform()`), 
are not available outside the function, so this will generate an error:
```{r}
zvalue
```

Here is a user-made function that calculates sample statistics (three digits as
default) and outputs these values in a named vector
```{r}
calculate_stats <- function(x, digits = 3) {
  n <- length(x)
  m <- mean(x)
  md <- median(x)
  s <- sd(x)
  minimum <- min(x)
  maximum <- max(x)
  out <- round(c(n, m, s, md, minimum, maximum), digits)
  names(out) <- c('n', 'mean', 'sd', 'median', 'min', 'max')
  out
}
```

Here is another user-made function:
```{r}
create_rsample <- function(n = 100, m = 100, s = 15) {
  # Creates a random sample of n observations with a specified mean (m) and 
  # standard deviation (s). The data is drawn from the normal distribution and
  # then linearly transformed to the specified mean and standard deviation.
  x <- rnorm(n)
  x <- (x - mean(x)) / sd(x)  # Sets mean(x) = 0 and sd(x) = 1
  x <- (x * s) + m  # Sets mean(x) = m and sd(x) = s
  x  # Return x
}
```

And here are both functions in action:
```{r}
sample1 <- create_rsample()
calculate_stats(sample1)
sample2 <- create_rsample(10, 50, 10)
calculate_stats(sample2)
```
If these were your first functions: Congratulations, you just wrote your first
computer programs!  

Note that it's a very good idea to comment your code carefully, including your 
functions (as I did above for `create_rsample()`). The time spend on commenting
your code is usually much less then the time you will need to understand your 
uncommented code when revisited on another day. 
  
\  


### Exercises
This code creates an integer vector randomly drawn with replacement from the set
of integers between 1 and 10. 
```{r}
set.seed(123)
randint <- sample(seq(1, 10), 10, replace = TRUE)
randint
```

Create a function that takes a vector created in this way as input and returns
a vector with three statistics: (i) the number of unique integers, (ii) their mean
value (excluding the duplicates), and (iii) the mean of the non-included integers
between 1 and 10.  
If the `randint` vector defined above would be given as input to this function,
something like this would be returned:
```{r, echo = FALSE}
do_exc1 <- function(x, n) {
  uni <- unique(x)
  a <- length(uni)
  b <- mean(uni)
  nonin <- setdiff(1:n, uni)
  c <- mean(nonin)
  out <- c(a, b, c)
  names(out) <- c('#included', 'mean-included', 'mean-not-included')
  out
}
do_exc1(randint, 10)
```

\ 

## Script 1: Randomized experiment (continued)
Click
<a href="script1_4.R" target="_blank">here</a>
for Script1 as we left it after the last seminar. 

Today we will rewrite the script to include potential outcomes on math scores, we will then 
calculate (unobservable) Single-unit causal effects and Average treatment effect (ATE), and 
compare these with estimates based om observed data.

\  



\  
\  

