---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX001
This seminar takes a first look at R and R-Studio. We'll
start with the Console, where you may interact directly with R. We will then
move to the text editor and start working on a simulation of a randomized experiment 
that we will continue to work on in the following seminars. 

\  

## Getting started with R and R-studio
Make sure you have R (the statistical software) and R-studio (an interface
to R) installed on you laptop before this seminar. You may run R without R-studio
(but not vice versa), but I strongly recommend that you use R-Studio; it makes 
working with R much easier.

### The Console
You may interact directly with R through the Console (typically the 
bottom left panel in R-studio). Here are some examples, just type the commands 
and hit Enter.

#### Use as calculator
$27 - 13$
```{r}
27 - 13
```

$(5+3)8^{1/3}$ 
```{r}
(5 + 3) * 8^(1/3)
```

$7!$ 
```{r}
factorial(7)
```

${7 \choose 2}$
```{r}
choose(7, 2)
```

\  

#### Get help (this opens a page in the Help panel of R-studio) 
```{r, eval = FALSE}
help(choose)
```

\  

#### Define data structures
Store a set of numbers in a vector called x
```{r}
x <- c(1, 7, 7, 3, 23, -4, 2^3)
```

Note that x is not displayed in the Console unless called specifically
```{r}
x  # This is shorthand for print(x)
```

Randomly select (with replacement) 10 integers 
between 1 and 20, and store them in a vector called y
```{r}
y <- sample(seq(1, 20), size = 10, replace = TRUE)
y
```

Repeat 'red', 'green' four times and store them in a vector called condition
```{r}
condition <- rep(c('red', 'green'), 4)
condition
``` 

\  

#### Check objects
The Console is useful for checking properties of R objects, for example, the 
number of elements in a vector, or its first (or last) values.
```{r}
a <- rep(month.abb, 5)  # help(month.abb) answers your question
length(a)  # Number of elements in vector a
head(a, 7)  # The first 7 values of vector a
tail(a, 5)  # The last 5 values of vector a
```

A common mistake is to try to do things on the wrong type of data, such as
trying to do math on a character (or string) vector.
Experienced R-users often use `str()` to remind them about what
type of data they are working with (some of this information is also displayed in
the Environment panel of R-studio). 
```{r}
a + 4
str(a)
```


\  

#### Asking the Console to produce plots works fine:
```{r}
random_data <- rnorm(10^4)
hist(random_data)
```

\  

#### Exercises 
1. Calculate $\left|\sqrt{10}/(\pi-4)\right|$. Remember that 
Google is your best friend. Use it to find out how R implements the 
square-root function, the absolute function, and the constant $\pi$.
The expression equals:
```{r, echo = FALSE}
abs(sqrt(10) / (pi - 4))
```
2. Use R's help function to find out what the `combn()´ function does. If the help 
page doesn't make sense, use Google to find out (R's help-pages take some
time to get used to).
```{r}
z <- combn(7, 2)
```
3. What's the difference between the `sort()` and the `order()` function?
Try this to find out.
```{r}
my_data <- sample(10)
a <- sort(my_data)
b <- order(my_data)
```

\  

### Script files
You will do most of your typing in the script window (typically located above
the Console in R-Studio). Here you write blocks of code in a script file
(select File/New File/R Script or press Ctrl+Shft+N in  R-Studio to open a new 
script). The code in the script is not executed unless you ask for it, 
for example by pressing Ctrl-R. If you just want to run a part of the script, 
selected this part and then hit Ctrl-R. If you just want to run a single line, 
just put the cursor on the line and hit Ctrl-R. 

\  

### Script1: Randomized experiment
Click
<a href="script1_0.R" target="_blank">here</a>
for a script file (Script1), to be completed it in the following seminars.  

The simulated scenario is a simple randomized experiment on the effect of extra 
music lessons during a semester on sixth-graders performance on a math test taken
at the end of the semester.  
(loosely inspired by Schellenberg, E.G. (2004). Music lessons enhance IQ. 
*Psychological Science*, 15, 511-514.) 

Key variables:

1. Independent variable: At the start of grade 6, all children at a school
are randomly assigned to receive either extra music lessons (treatment) or no extra
music lessons (control).    
2. Background (or pre-treatment) variable: Socioeconomic status, defined based 
on residential area. The children are either from a high-income
or a low-income residential area. 
3. Outcome variable: Score on a math test taken at the end of the semester.

Today, we will just set the scene:

1. Decide sample size and define a variable with id-numbers for each 
simulated participant
2. Simulate values on the pre-treatment variable (socio-economic status), using
the `sample()` function.
3. Sanity check using the `table()` function.

#### Exercises
In the following seminar (REX002), we will randomly assign children to experimental
conditions. Please try it beforehand. Note that there are different
approaches to random assignment, and for each, there are different ways to 
implement them in R-code.


\  
\  



