---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```

\  
\  


# Notes REX002
This seminar will first review the fundamentals of how R deals with data.
After that, we will continue our work on Script 1, today with random assignment
of our simulated participants to treatment and control conditions.

## How R deals with data
Please spend some time on this topic. It may seem abstract, even boring, but 
it is worthwhile studying. You don't have to memorize all this, just play around 
with the examples below to get a feel of data storage in R.  

One lesson to learn is that R sometimes 'help' you to make sense. This can indeed
be helpful, if you know what you are doing, but may cause strange errors if you
don't; errors that may hard to fix if you lack basic understanding of
how R deals with data.  

In R, data are contained in data structures. We will mainly work with two types
of data structures: Atomic vectors and data frames, so let's concentrate on 
them (and save lists, matrices and arrays for later).

### Vectors
Atomic vectors, or vectors for short, are 1-dimensional collections of elements
of the same type. The most important vector types are

1. Integer vectors
2. Numeric (or double) vectors
3. Character vectors
4. Logical vectors

Note that R treats a single data element as a vector with one element (i.e., R
has no scalars).  
When you work with numbers in R, they will most often be be defined as
numeric vectors. If you input a non-fractional number, R will treat it
as a numeric vector unless you tell it otherwise.
```{r}
a <- 4
str(a)  
b <- 4L  # you need to add an L to make it an integer
str(b)
```

Character vectors are defined using single (or double) quotes:
```{r}
x <- 'four'
str(x)
question <- "Did you hear a tone?"
str(question)
```

Logicals have two values, TRUE or FALSE. Typically, you get a logical
data element by making a statement:
```{r}
x <- 4 > 5  # I say: "4 is greater than 5". R answers: "FALSE"
x
str(x)
```

Sometimes you can and sometimes you can't do operations on elements of different
types. Always be careful when you can, to make sure you get what you expect.  
```{r}
# Adding numeric and integer works fine, output is numeric
x <- 4 + 4L  
x
str(x)

# Adding numeric and character doesn't work
y <- 4 + 'four'

# Adding a numeric and logical element works (surprisingly); useful for recodings of 
# variables, more on this later on ...
z <- 4 + TRUE
z
str(z)  # output is numeric

```

\ 

You may combine several data elements (i.e., one-element vectors) in a vector 
using the concatenation function `c()`.
```{r}
x <- c(1, 2, 3, 5, 8, 13, 21)
x

y <- c('1', '2', '3')
y

z <- c(TRUE, 2 != 5, FALSE)  # (2 != 5) is the same as !(2 == 5)
z

w <- c(1, '2', 3)  # Here R simply assumes that you want a character vector
w
```

#### Factors
Factors stores categorical data, and can contain only predefined values. For example, 
you may have a numeric vector ```conditions```  with three values, 1, 2, 3, 
representing treatment group, placebo group, and no-placebo-control group.

```{r}
conditions <- rep(c(1, 2, 3), 3) 
str(conds)
```
You may use this vector to define a new vector that is a factor:
```{r}
conds_factor <- factor(conditionss, levels = c(1, 2, 3), 
                       labels = c('treatment', 'placebo', 'no-placebo'))
str(conds_factor)
conds_factor
```
A factor is in fact an integer vector with specific attributes that (i) defines
the set of allowed values (levels) and (ii) changes its behavior compared to an
ordinary integer vector. For example, you cannot do arithmetic on a factor. If
you try, missing data, ```NA```, is generated. 
```{r}
conds_factor - 1  # No error message, but a warning that NA's has been generated
conds - 1  # Arithmetic on the numeric vector of course works fine
```
It may therefore be a good idea to keep the original numeric (or integer) vector
and define the factor as a new variable (as done in the example above)


#### Names of vector elements
You may name elements of a vector to help you remember what they stand for.
Names are attributes (meta-data) attached to vectors:
```{r}
iq <- rnorm(500, 100, 15)
iq_stats <- c(mean(iq), sd(iq), max(iq))
iq_stats
names(iq_stats) <- c('mean', 'sd', 'maxímum')
iq_stats
```


### Data frames
A data frame is a 2-dimensional data structure, that combines vectors of equal
length in rows and columns. Each vector (column) must contain the same type of
elements,but different columns may contain different types of elements. The code
below creates three vectors and then combines them into a data frame using the
```data.frame()``` function.
```{r}
id <- seq(1, 200)
group <- sample(c('male', 'female'), 200, replace = TRUE)  # character vector
performance <- rnorm(200, 100, 20)
my_data <- data.frame(id, group, performance)
```

```my_data``` is a data frame with 200 observations in 3 columns. Note that the
second column ```group``` contains a factor (not a character vector as one
would have expected). This is an example where R's helpfulness may cause problems. 
The intention is good, because you often want character vectors to be factors.
But assume that your data were based on a gender question with a third alternative,
'other', that no-one in your sample selected. R would assume that there were only 
two alternatives:
```{r}
table(my_data$group)
```
You have to fix this manually, using ```levels()``` (use $ to select a column
from a data frame. More on subsetting in REX003):
```{r}
levels(my_data$group) <- c('male', 'female', 'other')
table(my_data$group)
```
(The may use the argument ```stringAsFactors = FALSE``` when calling the 
```data.frame()``` function to keep character vectors as they are.) 

### Exercises
1. What's happening here? Explain line by line.
```{r}
set.seed(123)
new_data <- rnorm(100)
above_average <- 1 * (new_data   > 0)
group <- factor(above_average, levels = c(0, 1), labels = c('blue', 'green'))
my_frame <- data.frame(new_data, above_average, group)
str(my_frame)
```



## Script 1: Randomized experiment (continued)
Click
<a href="script1_1.R" target="_blank">here to view</a>
Script1 as we left it after the last seminar (if you missed it, please copy and
paste the content into a new script file). 

Today we will do the following:

1. Combine created variables in a data frame using the function ```data.frame()```
2. Randomly assign to treatment conditions, using the function call ```sample(c(0, 1), 
replace = TRUE, size = n)```, where ```size = n``` refer to sample size. That is,
randomization using the Bernoulli trial method.
3. Do some checks of balance between treatment and control group in socio-economic
status (SES). If random assignment works, the groups should be approximately equal 
in SES (indeed on all variables, measured and unmeasured!)

### Exercises
In the next seminar (REX003) we will simulate potential and observed outcomes.
Please give it a try already now.  
Remember from Method1, that the observed outcome
, $Y_{obs}$ is given by  

\ 
$Y_{obs} = DY_1 + (1-D)Y_0$,   
\
where $Y_0$ is the potential outcome if untreated,$Y_1$ the potential outcome 
if treated, $D = 1$ for those that received treatment, and $D = 0$
for those who didn't (assume, for now, that assigned and received
treatments are identical).

\  
\  
