---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX105
Today just a little more on functions, especially on those that take functions as 
arguments (so called functionals). There are many useful functionals in R, including 
functions in the apply family. Below I will just give a few examples. (Google 
'R apply family' to find numerous examples and tutorials.)

\  
\  

### Functions as arguments
Some functions accept functions as arguments. We have already used some of these, 
for example `apply()` that applies a function to the rows or columns 
of a homogeneous data frame (a data frame with only one type of elements, 
i.e., a matrix). Here is another example:

```{r}
# Simulate responses on 10 questionnaire items, each scored 1, 2, ..., 5
set.seed(123)
d <- data.frame(replicate(10, sample(1:5, 100, replace = TRUE)))

# Means across rows
meanrows <- apply(d, MARGIN = 1, FUN = mean) 
summary(meanrows)
# Means across columns
meancols <- apply(d, MARGIN = 2, FUN = mean) 
meancols
```

\  
\  

### Anonomous functions
Anonymous functions are functions without name. Typically they are performing 
simple tasks that can be defined in a single row of code. Anonymous functions are
frequently used in functionals. Here is an example using the data generated above. 
The anonymous function `function(x) sum((x - mean(x)^2))` calculates the sum of 
squares of its input, which in this case is the columns of d (as I set MARGINS = 2):
```{r}
# Calculate sum of squares across columns of d
sscols <- apply(d, MARGIN = 2, FUN = function(x) sum((x - mean(x))^2)) 
sscols
```

\  
\  

### lapply
There are a whole set of functions in the apply family. One of the more useful is 
`lapply()` that does something on each element of the input and returns the result 
in a list. Here is an example on how to use it on a data frame containing columns 
of different types.
```{r}
# iris is a datset preloaded in R. This checks the class of each column
lapply(iris, FUN = class)
# This counts missing values in each column
lapply(iris, function(x) sum(is.na(x)))
```
The return is a list. If you prefer to have it as a vector, try `sapply()`
(s for simplify)
```{r}
sapply(iris, class)
sapply(iris, function(x) sum(is.na(x)))
```

\  
\  

### tapply
As a final example of functionals, let's have a look at the function `tapply()`. 
This function allows you to apply a function on subsets of your data. Here is a 
simple example calculating the median sepal length for each species in the
`iris` dataset.
```{r}
tapply(iris$Sepal.Length, INDEX = list(iris$Species), FUN = median, na.rm = TRUE)
```
Note that the second argument (INDEX) has to be a list. It may be a list of several 
variables by which to split the analysis:
```{r}
# Create a new dataframe from the iris dataset
d <- iris
# Add dichotomous variable, distinguishing narrow and wide sepals.
d$sepal_width_dicho <- factor(1 * (d$Sepal.Width > median(d$Sepal.Width)), 
                              levels = c(0, 1), labels = c('narrow', 'wide'))

# Calculate median sepal length for each species and sepal width category
tapply(d$Sepal.Length, INDEX = list(d$Species, d$sepal_width_dicho), FUN = median)
```


\  
\  
\ 
\  




### Script2: Analysing data from a psychoacoustic experiment
Click
<a href="script2_4.R" target="_blank">here</a>
for the script file (Script2) as we left after last time.  
(Data sets:  <a href="./stat1/threshold_data.txt" target="_blank">threshold data</a>, and 
<a href="./stat1/background_data.txt" target="_blank">background data</a>. See 
<a href="./stat1/codebooks.txt" target="_blank">codebooks.txt</a>
for details on the variables).

Today some bootstrapping  of differences between group medians. Maybe there will 
be time also for some robust regression (least absolute value regression). 

#### Exercises
Google to find out various ways to implement least absolute value regression.  


\  
\  



