---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX104
Today some tips on how extract and repackage results from analysis stored as 
R objects. User-made functions can be helpful for customize outputs from 
standard analysis (like the t-test or linear regression). Then we will continue 
our work with analyzing data from the psychoacoustic experiment (Script 2).


### Outputs are objects
Everything in R is an object. This is true also for outputs of statistical tests.
Here is a t-test of the difference in mean sleep improvement between two groups 
taking two different sleeping pills (1 and 2). 
```{r}
# Data from R's set of preloaded data
d <- sleep
# Just recoding, so group variabel is dummy coded: group 1 = 0, group2 = 1
d$group <- as.integer(d$group) - 1
# t-test, I prefer the Welch test (var.equal = FALSE), but here I use the standard
# t-test (var.equal = FALSE), to compare with the output of the dummy variable 
# regression below
t_out <- t.test(d$extra[d$group == 1], d$extra[d$group == 0], var.equal = TRUE)
t_out
```

I stored the output in `t_out`. Use `attributes()` to see the names of the 
various data structures stored in `t_out`; use `$` to extract selected 
output. 
```{r}
attributes(t_out)
t_out$conf.int
```

This code combines selected output in a vector called `t_result`. Use 
`names()` to remember what each element represents.
```{r}
# Difference between group means is not stored in t_out, so calculate below
mean_diff <- mean(d$extra[d$group == 1]) - mean(d$extra[d$group == 0])  
t_result <- c(mean_diff, t_out$conf.int, t_out$p.value)
names(t_result) <- c('mean-diff', 'ci95%lo', 'ci95%hi', 'p-value')
round(t_result, 3)  # Print to screen with three decimals
```

\  

Here is a regression analysis using the same data. The fitted model is 
$\hat{y} = b_0 + b_1D$,  
where $D$ is a dummy coded variable of group: group1 = 1, group2 = 0.  
```{r}
# Just to recode to Group1 = 1, Group2 = 0, before the regression analysis, lm()
lm_out <- lm(d$extra ~ d$group)
lm_out
attributes(lm_out)
```

```{r}
# Confidence intervals are not in lm_out, calculated below
b_ci <- confint(lm_out)[2, ]  # Confidence interval about coefficient for group-dummy
# A bit hard to find the p-value of the regr. coefficient, but here it is:
pvalue <- summary(lm_out)$coefficients[2, 4]
# Packaging the regression results in a numeric vector
lm_result <- c(lm_out$coefficients[2], b_ci, pvalue)
names(lm_result) <- c('regr.coeff', 'ci95%lo', 'ci95%hi', 'p-value')
round(lm_result, 3)  # Print to screen with three decimals
round(t_result, 3)  # ... and t-test results again for comparison
```


\  

#### Output as input

Output stored as an object may be used as input to other functions. Here is a plot 
of the relationship between extra sleep and the dummy coded group 
variable. `lm_result` was used as input to the function `abline()` to draw the 
regression line.

```{r, fig.width = 3}
plot(d$group, d$extra, xlab = 'Group \n(0 = group1, 1 = group2)',  
     ylab = 'Extra sleep', xlim = c(-0.1, 1.1), axes = FALSE)
axis(1, at = c(0, 1), tck = 0.02)
axis(2, las = 1, tck = 0.03)
lm_out <- lm(d$extra ~ d$group)
abline(lm_out)
```
 
 

\  
\  
\ 


### Script2: Analysing data from a psychoacoustic experiment
Click
<a href="script2_3.R" target="_blank">here</a>
for the script file (Script2) as we left after last time.  
(Data sets:  <a href="./stat1/threshold_data.txt" target="_blank">threshold data</a>, and 
<a href="./stat1/background_data.txt" target="_blank">background data</a>. See 
<a href="./stat1/codebooks.txt" target="_blank">codebooks.txt</a>
for details on the variables).


Today we'll sum up our group comparisons based on arithmetic means, and then do 
similar analysis of differences between group medians (robust to influences from 
the outliers in the data). Finally, we will then do test the relationships between 
auditory abilities and age using regression techniques. 

\ 

#### Exercises
How would you evaluate the observed differences between group medians of 
auditory thresholds?  


\  
\  



