---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX103
Today a little about recoding in R. There are many situations in which one wants 
to recode a variable. Below some examples of how to do this.

### Factors
If you have categorical data coded with integers, its easy to turn them into factors. This is not really a recoding, but rather naming your categories. We have used 
factors earlier, here an illustration that demonstrates the difference between 
unordered and ordered factors.
```{r}
# Simulate questionnaire data on residential area.
area <- sample(x = c(1, 2, 3), size = 500, replace = TRUE)
table(area)
area <- factor(area, levels = c(1, 2, 3), 
                         labels = c('rural', 'small-town', 'city'))
table(area)
```

\  


\  
\  
\  
\ 


### Script2: Analysing data from a psychoacoustic experiment
Click
<a href="script2_2.R" target="_blank">here</a>
for the script file (Script2) as we left after last time.  
(Data sets:  <a href="./stat1/threshold_data.txt" target="_blank">threshold data</a>, and 
<a href="./stat1/background_data.txt" target="_blank">background data</a>. See 
<a href="./stat1/codebooks.txt" target="_blank">codebooks.txt</a>
for details on the variables).


Today we will do more detailed group comparisons, including statistical tests and calculations of confidence intervals around between-group differences. We may also have time to make some preliminary checks of the relationship between auditory abilities and age using regression techniques. 

\ 

#### Exercises
How would you evaluate statistically the observed differences in auditory abilities between the blind group and the sighted groups? 


\  
\  



