---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX101
This seminar deals with importing data to R. There are many ways to do this;
we'll use the function `read.table()`. Once this has been mastered, we will import 
some real data and start working on a full data analysis of data from an 
experiment (Script2, to be completed in the following REX seminars).

\  

### Set the working directory
The way to start a data analysis project is to first create a directory 
dedicated to project. This directory will contain data files, code books, your 
R-scripts, and, later on, output from your analyses (e.g., figures).   

The code below checks the current directory, and then changes it to a directory
containing the data set to be analysed. The functions `list.dirs()` and 
`list.files()` do what their names imply (as function names should do!).

```{r }
getwd()
list.dirs()  # List directories in the current directory
setwd("./stat1")  # Move to new directory ( . means current directory)
getwd()  # Check that we are in the new directory 
list.files()  # List files in the (new) current directory
```

\  

### Import data
We will use the function `read.table()` to import data in R. The function imports 
a text file with data ordered in columns. The output is a data frame. 
The function's first argument is the path to the text file. If you just give a 
file name, R will look in the current directory (if it is not there R will 
through you an error message). If you want to import a dataset located in 
another directory you have to specify the path (or change R's current 
directory using `setwd()`).  

In the example below, the to be imported text-file has variable names in the first 
line and the columns are separated by commas, hence the arguments 
`header = TRUE` and `sep = ","`. The data is from 
<a href="howell_table_15_1.txt" target="_blank">Howell's Table 15.1</a> 

```{r }
d <- read.table("howell_table_15_1.txt", header = TRUE, sep = ",")
```


\  

#### Initial data screening
Below some initial checks of the imported data. Note from the output of `str(d)` 
that R imports variables with decimal values as type `num`, variables with integers
as type `int`, and that variables with characters (strings) are imported as
factors.
```{r }
# Check that d is a data frame. 
str(d)
# List column names
colnames(d)
# Check that the data looks alright, by looking at the first few lines
head(d)
# ... and at the last few lines
tail(d)
# Some summary statistics
summary(d)
# Here I add two NA's, just for illustration
d[c(7, 31), c(5, 6)] <- NA
# Summary statistics of selected variables, note the NA's
summary(d[, c(5, 6)])
sum(is.na(d))  # This counts the NA's in the data frame
```

\  
\  

### Script2: Analysing data from a psychoacoustic experiment
Click
<a href="script2_0.R" target="_blank">here</a>
for a script file (Script2), to be completed it in the following seminars.  

The data is from a listening experiment involving blind and sighted 
listeners, conducted at the Gösta Ekman Laboratory in 2016. The experiment measured 
auditory thresholds for abilities potentially involved in human echolocation.  

The data is stored in two files. The first file contains 
<a href="./stat1/threshold_data.txt" target="_blank">threshold data</a>, the 
second file contains 
<a href="./stat1/background_data.txt" target="_blank">background data</a> 
(age, sex, ...). Please see the file 
<a href="./stat1/codebooks.txt" target="_blank">codebooks.txt</a>
for details on the variables stored in the data files.

The hypotheses of the experiment were straight forward: (1) Blind listeners were 
expected  to perform better (have lower thresholds) than the sighted listeners on 
all four conditions, but (2) age may moderate this effect, so differences between groups 
of blind and sighted listeners were expected to be larger for the age-matched than 
the young group of sighted listeners.  

Today we will just import the data files and do some preliminary screening tests 
to make sure the data was imported correctly.

\ 

#### Exercises
Try to do some initial between-group comparisons of thresholds, separately for 
each of the four experimental conditions.

\  
\  



