---
title: False
author: "Mats E. Nilsson"
---

```{r setup, cache = FALSE, echo = FALSE}
# THis will allow error messages to be converted to HTML
knitr::opts_chunk$set(error = TRUE)
```
\  
\


# Notes REX103
Today a little about recoding in R. There are many situations in which one wants 
to recode a variable. Below some examples of how to do this.

\  

### Factors 
If you have categorical data coded with integers, its easy to turn them into factors. This is not really recoding, but rather naming your categories. We have used factors earlier, here an example demonstrating the difference between
unordered and ordered factors.
```{r}
# Simulate data on residential area, where 1 = city, 2 = rural, and 
# 3 = town.
set.seed(123)
area <- sample(x = c(1, 2, 3), size = 500, replace = TRUE)
table(area)

# Make (unordered) factor variable
area_unordered <- factor(area, levels = c(1, 2, 3), 
                         labels = c('city', 'rural', 'town'))
table(area_unordered)
str(area_unordered)
```
Here the same but now ordering the factor levels with regard to population density
```{r}
# Make ordered factor variable (entered in the desired order)
area_ordered <- factor(area, levels = c(2, 3, 1), ordered = TRUE,
                         labels = c('rural', 'town', 'city'))
table(area_ordered)
str(area_ordered)
```

Ordered factors are useful when you want to display the categories in the right 
order, e.g., in a table as above. 

\  
\  

### Recoding using coercion of logicals to numericals

Variables can be recoded in many ways. Below I illustrate 
the method that I use the most. It uses R's automatic coercion to numericals when arithmetic operations are conducted on logicals.

```{r}
# Simulate some data (BMI in a population with overweight problems)
set.seed(123)
bmi <- rnorm(n = 1e4, mean = 29, sd = 4)

# Recode into four groups
bmi_group <- 1 * (bmi < 18) +
             2 * (bmi >= 18 & bmi < 25) +
             3 * (bmi >= 25 & bmi < 30) +
             4 * (bmi > 30)
table(bmi_group)

# Followed by making the new variabel a factor, if so desired
bmi_group <- factor(bmi_group, levels = c(1, 2, 3, 4), 
                    labels = c('underweight', 'normal', 'overweight', 'obese'))
table(bmi_group)
```

The same could be achieved in a shorter, but less transparent way using the function `cut()`:
```{r}
# Recode into four groups
bmi_group <- cut(bmi, breaks = c(-Inf, 18, 25, 30, +Inf), 
                 labels = c('underweight', 'normal', 'overweight', 'obese'))
table(bmi_group)
```
\  
\  
Coercion of logicals can be used for recodings based on several variables. Here 
an example with BMI and ill-health:
```{r}
# Simulate some data (BMI and ill-health in a population with overweight problems)
set.seed(123)
bmi <- rnorm(n = 1e4, mean = 29, sd = 4)
health <- sample(x = c('healthy', 'ill'), size = length(bmi), replace = TRUE,
                 prob = c(0.8, 0.2))

# Recode into three groups
obese_ill <- 1 * (bmi <= 30) +
             2 * (bmi > 30  &  health == 'healthy') +
             3 * (bmi > 30  &  health == 'ill')

obese_ill <- factor(obese_ill, levels = c(1, 2, 3), 
                    labels = c('non-obese', 'obese-healthy', 'obese-ill'))

table(obese_ill)
```
A slightly shorter, but less transparent code for recoding in k groups is to define k-1 groups, and let the remaining group take on the value 0 (the numeric coercion of FALSE):
```{r}
# Recode into three groups with only two definitions
obese_ill <- 1 * (bmi > 30  &  health == 'healthy') +
             2 * (bmi > 30  &  health == 'ill')
obese_ill <- factor(obese_ill, levels = c(0, 1, 2), 
                    labels = c('non-obese', 'obese-healthy', 'obese-ill'))
table(obese_ill)
```
  
\ 
\ 

### Combining vectors
Combining vectors is a common operation. For example, multiple items may be 
combined into a sum score or an indicator variable may be created based on the 
response pattern over a set of items. Combining vectors in a data frame is easy
in R:
```{r}
# Simulate data: five questionnaire items, scored from 1 to 5, stored in a data 
# frame called d. To complicate things, I also include a character vector in d
set.seed(123)
n = 999
d <- data.frame(paste(rep('id', n),as.character(1:n), sep = ''))
d[, 2:6] <- replicate(5, sample(x = 1:5, size = n, replace = TRUE))
colnames(d) <- c('id', 'item1', 'item2', 'item3', 'item4', 'item5')

# Add vector with sum score to the data frame d
d$sumscore = d$item1 + d$item2 + d$item3 + d$item4 + d$item5

# Add vector indicating if at least one of the items had value = 5
d$topscore <- 1 * (d$item1 == 5 | d$item2 == 5 | d$item3 == 5 | 
                    d$item4 == 5 | d$item5 == 5)
head(d)
```
This works OK with only five items. But with 20 or 30 or more items, the approach 
taken above would be unfeasible. Here is a smarter way, using the function 
`apply()`. It's input is a matrix, that is, a 2-dimensional object with one type of data. Therefore I cannot input the whole data frame, as it contains one 
character vector (d$id).
```{r}
# Use apply() to add vector with sum score
d$sum_apply = apply(d[, 2:6], 1, sum)

# Use apply() to add vector with top score (again using logi-to-num coercion) 
d$top_apply <- apply(d[, 2:6], 1, function(x) 1 * (5 %in% x))
head(d)
```

\ 
\  
\  



### Script2: Analysing data from a psychoacoustic experiment
Click
<a href="script2_2.R" target="_blank">here</a>
for the script file (Script2) as we left after last time.  
(Data sets:  <a href="./stat1/threshold_data.txt" target="_blank">threshold data</a>, and 
<a href="./stat1/background_data.txt" target="_blank">background data</a>. See 
<a href="./stat1/codebooks.txt" target="_blank">codebooks.txt</a>
for details on the variables).


Today we will do more detailed group comparisons, including statistical tests and calculations of confidence intervals around between-group differences. We may also have time to make some preliminary checks of the relationship between auditory abilities and age using regression techniques. 

\ 

#### Exercises
How would you evaluate statistically the observed differences in auditory abilities between the blind group and the sighted groups? 


\  
\  



